Measuring the double-drift illusion with hand tracking

### 2021-08-05 Manuscript.Rmd
edits suggested by Patrick

---
title: "Measuring the double-drift illusion with hand tracking"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document:
    reference_docx: template.docx
  html_document:
    df_print: paged
  pdf_document: default
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# Here we compare the tracking data with an earlier paper:
source('R/compare.R')
# And then some code for the re-tracing experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
# And finally, code to compare this data with Sirui's PSEs
source('R/time.R')

# we download the data:
OSFdata()
# this only downloads the data for the paper (well...)

# we pre-process the data:
getBoundedTrackingStandardizedSegments()
getSegmentDirections()
preProcessOnePass_V4()
```

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence its perceived location, making it appear to be slightly ahead of its true location, e.g. the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013]. These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Hogendoorn2020; @Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor has internal motion orthogonal to its path and is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015], and although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018]. These findings suggest that both the true and illusory positions are available in the brain.

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019]. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms. The perceived location conintues to move away from its physical position by up to several times the size of the gabor. The strength of the illusion, the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019]. In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig. 1) with 5 speeds of the internal motion. We analyze the angle of the tracking to determine if the manual response is susceptible to the illusion.

@@

Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses. Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet. The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus. Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].

@@

We will compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller, Patel, Faustin, Cavanagh, and Tse [-@Heller2021]. In this model, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$. That is, the deviation from the physical direction is given by:

@@

@@

```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data. **Left:** Individual trajectories. Note that there are only a few obvious discontinuities (highlighted in red). In general, the traced paths showed rough linearity. **Top right:** Average heading of the middle 1 _s_ of each pass. Distance from origin is time, angle is the heading angle. Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading. **Bottom right:** The strength of the illusion in degrees angle deviation from the physical path. In red are the average illusion strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants. Gray dashed line: prediction of the vector combination model of Equation 1 with $K=1$. Black line: fit with $K=0.81$ as in Cavanagh & Tse (2019), $K=0.81$. Purple: best fit to our data, $K=0.72$."}
plotBoundedTracking()
```

@@

The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it? In Figure 2 (bottom right), we compare mean illusion strength to that reported previously [@Cavanagh2019]. As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model (Eq. 1, in black). That is; tracking the illusion while observing it does not change the strength of the illusion.

@@

As before, we compare the strength of the illusion (see Fig 5, left), with an earlier study [@Cavanagh2019]. A line through the origin with slope 0.51 ($K$ from Eq. 1) predicts illusion strength fairly well in these data, which is weaker than the $K$=0.81 found by Cavanagh & Tse [-@Cavanagh2019] and lower than the $K$=0.72 found in Experiment 1 with online tracking. While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

@@

```{r fig5, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5:** Illusion strength and reset models. **Left:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities. Purple dots are the group means from 3 s passes, red from 4 s passes (2, 3 and 4 _cps_ internal speeds). Predictions from Eq. 1: Gray dashed line for $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval. **Right:** Model limits shown here come from a weighted bootstrapping across participants, lines are median parameter values, shaded areas the 95% confidence intervals of the limits. Gray dots are the reset points from all participants. While a spatial limit provides the best fit, neither a spatial nor a temporal limit explains the data well."}
#plotResetPoints()
#plotModels()
#plotBootstrappedFits()
plotIllusionStrength()
```

We test this with a series of simple models. First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength, expressed as either an angle (Fig 4, left: $a = atan2(X,Y)$) or a slope ($s = \frac{sin(a)}{cos(a)}$) to predict reset points.

If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a spatial limit ($L_x$) would predict the y-coordinate of the first reset, $y$:

@@

Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ corresponds to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus. If we fit the two simple models for each participant separately, by minimizing the mean square error (in isometric space, with absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit. We can also combine limits to see if they are complementary, including. Here, we see that for 4/9 participants (MSE and AIC) the best fit equals a spatial limit as the other limit is set to 0. Of the remaining participants 4 are best fit with a combined spatial and temporal limit (4/9) and one participant is best fit combining a vertical offset with a temporal limit. This indicates that in this data, a spatial limit explains the reset points best. However even the spatial limit does not explain the data very well.

To obtain population level values for the spatial and temporal limits, we resample reset points from each participant, using the smallest number of detected reset points that fall in the first quadrant, in the group (44), to ensure that each participant contributes the same number of reset points. Both models based on a spatial limit (eq. 2) and on a temporal limit (eq. 3) as well as a model combining these two were fit 1000 times. In the models with two limits there is a strong negative correlation between fitted spatial and temporal limits (r(998)=-0.985, p>.001) indicating parameter redundancy in the combined limit models. That means that the two limits probably explain the same parts of the data, making a two-limit model very hard to interpret. Hence, only the bootstrapped confidence intervals for the two single limit models are shown in Figure 5. Further, since temporal limits might be fit better with an exponential distribution (and minimizing errors amounts to using a normal distribution), we repeated fitting single limit models based on normal, exponential and gamma likelihood distributions for all participants, but found that of the best temporal models, only a few used an exponential distribution (3/9) and the rest a normal distribution (6/9). All model fits combined show that while the reset point data reported here do somewhat favour a spatial limit, they can not conclusively rule out other limits - most notably a temporal limit.

@@

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in this experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe also suggest that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. **However, “return” resets occurred at a fairly consistent location for each participant, which suggests that they are triggered by the stimulus dynamics and not by attentional distraction or eye movements.**

We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion. Few if any resets were detected during the online tracking, either because of the additional attentional requirements or because of the short duration (2 s), Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2. The resets were best explained by a limiting spatial offset from the true stimulus location, but do not exclude a temporal limit. These findings may either indicate there are multiple causes for spontaneous resets or that a re-tracing method is not suitable to resolve what causes this phenomenon. Either way, both a random time from stimulus onset until reset and too large a conflict between retinal and perceived position leading to resets remain possible.

>>>>

---
title: "Measuring the double-drift illusion with hand tracking"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document:
    reference_docx: template.docx
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# Here we compare the tracking data with an earlier paper:
source('R/compare.R')
# And then some code for the re-tracing experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
# And finally, code to compare this data with Sirui's PSEs
# source('R/time.R') # no longer used

# we download the data:
OSFdata()

# we pre-process the data:
getBoundedTrackingStandardizedSegments()
getSegmentDirections()
preProcessOnePass_V4()
# the rest of the paper should now compile without complaint
```

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence its perceived location, making it appear to be slightly ahead of its true location, e.g. the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013]. These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Hogendoorn2020; @Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor has internal motion orthogonal to its path and is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: immediate saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. In contrast, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018]. These findings suggest that both the true and illusory positions are available in the brain.

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019]. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves, as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms. The perceived location continues to move away from its physical position by up to several times the size of the gabor. The strength of the illusion, the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021]. In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig. 1) with 5 speeds of the internal motion. We analyze the angle of the tracking to determine if the manual response is susceptible to the illusion.

@@

Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses. Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet. The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus. Experiments were run in Python 2.7 with PsychoPy [@Peirce2019].

@@

We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021]. In this model, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$. That is, the deviation from the physical direction is given by:

@@

All analyses were done in R 3.6.1 [@Rproject]. All data, as well as scripts for the experiments and analyses are available on OSF: [https://osf.io/72ndu/](https://osf.io/72ndu/).

@@

```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data. **Left:** Individual trajectories. In general, the traced paths showed rough linearity. **Top right:** Average heading of the middle 1 _s_ of each pass. Distance from origin is time, angle is the heading angle. Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading. **Bottom right:** The strength of the illusion in degrees angle deviation from the physical path. In red are the average illusion strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants. Gray dashed line: prediction of the vector combination model of Equation 1 with $K=1$. Black line: $K=0.74$ as in Heller et al. (2021). Purple: best fit to our data, $K=0.72$."}
plotBoundedTracking()
```

@@

The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it? In Figure 2 (bottom right), we compare mean illusion strength to that reported previously [@Heller2021]. As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model fitted to all their data (Eq. 1, in black). That is; tracking the illusion while observing it does not change the strength of the illusion.

@@

As before, we compare the strength of the illusion (see Fig 5, left), with an earlier study [@Heller2021]. A line through the origin with slope 0.51 ($K$ from Eq. 1) predicts illusion strength fairly well in these data, which is weaker than the $K=0.74$ found by Heller et al. [-@Heller2021] and lower than the $K=0.72$ found in Experiment 1 with online tracking. While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

@@

```{r fig5, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5:** Illusion strength and reset models. **Left:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities. Purple dots are the group means from 3 s passes, red from 4 s passes (2, 3 and 4 _cps_ internal speeds). Predictions from Eq. 1: Gray dashed line for $K=1$, black line; $K=0.74$, light blue line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval. **Right:** Model limits shown here come from a weighted bootstrapping across participants, lines are median parameter values, shaded areas the 95% confidence intervals of the limits. Gray dots are the reset points from all participants. While a spatial limit provides the better fit, neither a spatial nor a temporal limit explains the data well."}
#plotResetPoints()
#plotModels()
#plotBootstrappedFits()
plotIllusionStrength()
```

We test this with a series of simple models. First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength based on a point earlier in the trajectory, expressed as either an angle (Fig 4, left: $a = \arctan{(X/Y)}$) or a slope ($s = sin(a)/cos(a)$) to predict reset points.

If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a spatial limit ($L_x$) would equal the x-coordinate and predict the y-coordinate of the first reset, $y$:

@@

Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ corresponds to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus. If we fit the two simple models for each participant separately, by minimizing the mean square error (in isometric space, with absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit. This indicates that in this data, a spatial limit explains the reset points best, but neither limit explains the data very well.

To obtain population level values for the spatial and temporal limits, we resample reset points from each participant, using the smallest number of detected reset points that fall in the first quadrant, in the group (44), to ensure that each participant contributes the same number of reset points. Both models based on a spatial limit (eq. 2) and on a temporal limit (eq. 3)  were fit 1000 times. We explored other modeling approaches (e.g. based on various distributions of likelihoods or combining limits) but these show the same pattern of results: spatial limits are somewhat favoured, but temporal limits can not be excluded (see OSF repository: [https://osf.io/72ndu/](https://osf.io/72ndu/)).

@@

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in this experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe also suggest that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the modestly better performance of the spatial limit model here suggests that spontaneous resets are not randomly distributed - as they might be if eye-movements or shifts of attention are the only way to elicit resets.

We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion. Few if any resets were detected during the online tracking, either because of the additional attentional requirements or because of the short duration (2 s). Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2. The resets were best explained by a limiting spatial offset from the true stimulus location, but do not exclude a temporal limit. These findings may either indicate there are multiple causes for spontaneous resets or that a re-tracing method is not suitable to resolve what causes this phenomenon. Either way, both some amount of time from stimulus onset until reset and too large a conflict between retinal and perceived position leading to resets remain possible.
